---
title: The impact of parental educational attainment on mathematics achievement among high school students

author: "Eva Y."
date: "2023-04-14"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

The purpose of this analysis is to investigate whether parental educational attainment has an impact on high school students' mathematics achievement measured by math scores of a standardized test.

## Load data and dependencies

```{r load dependencies}
library(data.table)
library(ggplot2)
```

```{r load data}
math_dat <- fread("../data/exams.csv", sep = ",")
```

## Data exploration

We're going to explore the data to gain insights into its features (e.g., number of columns and rows, data types). 

```{r explore data}
# str() gives me a general idea of the data (e.g., it has 1000 rows and 8 columns)
math_dat[, str(.SD)]
```

```{r check out top rows}
# print top rows
math_dat[, head(.SD)]
```

### Metadata

Let's refer to the data source and its metadata. This dataset is downloaded from [here](https://www.kaggle.com/datasets/rkiattisak/student-performance-in-mathematics). Based on the metadata, it contains the following features so this seems to line up with what we observed so far in terms of data types (e.g., math score is an integer data type).

* Gender: The gender of the student (male/female)
* Race/ethnicity: The student's racial or ethnic background (Asian, African-American, Hispanic, etc.)
* Parental level of education: The highest level of education attained by the student's parent(s) or guardian(s)
* Lunch: Whether the student receives free or reduced-price lunch (yes/no)
* Test preparation course: Whether the student completed a test preparation course (yes/no)
* Math score: The student's score on a standardized mathematics test
* Reading score: The student's score on a standardized reading test
* Writing score: The student's score on a standardized writing test

For the purpose of this assignment, we will focus on the parental level of education and math score columns since we are interested to know whether parental educational attainment has an impact on math scores among high school students.

### Descriptive statistics

Another way to explore the data is to look at descriptive statistics. For example, this will give us insight into the mean test scores for the different parental level of education.

```{r descriptive stats}
# get the descriptive stats of each column
math_dat[, as.list(summary(`math score`)), by = `parental level of education`]
```

```{r standard dev}
# calculate sd for each parental level of education
math_dat[, sd(`math score`), by = `parental level of education`]
```

```{r frequency of students for different parental level of education}
math_dat[, .N, by = `parental level of education`]
```

### Missing values

Before moving on, we should also check for missing values in the dataset.

```{r check for missing values}
# count number of missing values in each column
math_dat[, lapply(.SD, function(x) sum(is.na(x))), .SDcols = 1:8]
```

There are no missing values in this dataset so we don't have to conduct any pre-processing to account for missing values.

### Data pre-processing

There are a few things we should clean up before moving on to visualization and analysis. First, let's change how the columns are named.

```{r clean up column names}
# replace space with underscore
new_colnames <- gsub(" |/", "_", names(math_dat))

# replace column names in the data with new_colnames
setnames(math_dat, names(math_dat), new_colnames)

# let's check out the new column names
names(math_dat)
```

Let's change parental level of education to a categorical variable.

```{r create factors}
# change parental_level_of_education to factors
educ_levels <- c("some high school", "high school", "some college", "associate's degree", "bachelor's degree", "master's degree")

math_dat[, parental_level_of_education := factor(parental_level_of_education, levels = educ_levels)]

# check if this has changed
str(math_dat)
```

From here onwards, we will also work with a subset of the data since we are only focusing on parental level of education and math score.

```{r subset data}
education_math <- math_dat[, .(parental_level_of_education, math_score)]
```

## Data visualization

We will visualize the math scores for each category of parents' education level using a boxplot. 

```{r visualize data using boxplot}
# plot boxplot
ggplot(education_math, aes(x = parental_level_of_education, y = math_score, fill = parental_level_of_education)) + 
  geom_boxplot() +
  theme_bw() +
  theme(legend.position="none", 
        axis.title.x = element_text(size=14, face="bold"),
        axis.title.y = element_text(size=14, face="bold")) +
  labs(x = "Parents' Level of Education", y = "Math Score")
```

```{r save educ math plot, echo=F, eval=F}
ggsave("../plots/educ_math.png", width = 8, height = 6)
```

## One-way ANOVA

Using a one-way ANOVA, we will determine whether parents' level of education has a significant impact on mean math scores of high school students.

```{r anova analysis}
# compute the ANOVA
math.aov <- aov(math_score ~ parental_level_of_education, data = education_math)

# summary of the analysis
summary(math.aov)
```

Based on this output, parents' level of education has a significant impact on the math scores of high school students. 

Now, we need to perform a post-hoc test to determine pairwise comparisons. For the one-way ANOVA, we will be using the Tukey's HSD post-hoc test.

```{r tukey pairwise comparison}
# run tukey post-hoc test
TukeyHSD(math.aov, conf.level = .95)
```

The Tukey's HSD post-host test demonstrated significant differences in math scores between the following pairs: 

* some college-some high school
* associate's degree-some high school
* bachelor's degree-some high school
* master's degree-some high school
* associate's degree-high school. 

No significant differences in math scores were found between the following pairs: 

* high school-some high school
* some college-high school
* bachelor's degree-high school
* master's degree-high school
* associate's degree-some college
* bachelor's degree-some college
* master's degree-some college
* bachelor's degree-associate's degree
* master's degree-associate's degree
* master's degree-bachelor's degree.

## Intepretation

* A one-way ANOVA was conducted to examine the impact of parents' level of education on high school students' math scores in a standardized test. 

* The ANOVA showed that the impact of parents' level of education on high school students' math scores was significant (*F*(5, 994) = 5.511, *p* < .05).

* A Tukey's HSD post-hoc test revealed that the mean math score of students whose parents completed some high school was significantly different than that of students whose parents completed some college, associate's degrees, bachelor's degrees, and master's degrees (*p* < .05). 

* The post-hoc test also showed that the mean math score of students whose parents completed high school was significantly different than that of students whose parents completed associate's degrees (*p* < .05). 

* No significant differences in math scores were found between high school and some high school, some college and high school, bachelor's degree and high school, master's degree and high school, associate's degree and some college, bachelor's degree and some college, master's degree and some college, bachelor's degree and associate's degree, master's degree and associate's degree, and master's degree and bachelor's degree (*p* > .05).

The end.
